Forecasting future oil production in Norway and the UK: a general improved methodology

### 2014-07-10 Publications/Forecasting  Future Oil Production/peak_oil_5.tex
Latest version.

%% LyX 2.0.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{article}
\usepackage{mathpazo}
\renewcommand{\ttdefault}{mathpazo}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{float}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{esint}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 1},backref=section,colorlinks=false]
 {hyperref}
\usepackage{breakurl}

@@

The idea behind the methodology is to model the future aggregate oil
production of a country by studying the production dynamics of its
individual constituents, the oil fields. The main benefit of this
approach, compared to working directly with aggregate production data,
is the possibility to forecast non-trivial oil production profiles
arising from the combination of all the individual field
dynamics. This ability reflects directly that the total
production is indeed the sum of the contributions of the individual
oil fields in production. Thus modelling at the level of each field
reflects more closely the reality and is likely to be over-performing
and more reliable, as we show below.
In order to implement our approach, one must be able 1) to \textbf{extend
the oil production of each individual field} into the future and 2)
to \textbf{extrapolate the rate of discoveries} of new oil fields.

@@

The first step to predicting the future oil production of a country
is to extrapolate the future production of existing fields and to
estimate the error on this extrapolation. The data of the fields developed
in the past shows a repeating asymmetric pattern. A good example is
the Oseberg field shown in figure \ref{regular}, with a quick ramp
up once the field is being developed, and then a peak or plateau before
the oil field production starts decaying. The decay can take many different shapes
and is governed by a variety of geological and economical factors.
The goal of the fitting procedure is to capture as much as possible the impact of
these different factors.

@@

The resulting plot shown in figure \ref{rate_of_discovery} pictures
the dynamics: giant oil fields have mostly been found while the discovery
process for dwarf fields is still ongoing. The logistic growth curve
fit to the giant discoveries is well constrained, however the fit
for the dwarfs is poorly constrained. As can be seen in figure \ref{rate_of_discovery},
the carrying capacity $K$ of the logistic growth model is not well
constrained by the available data. A large spectrum of values for
$K$ can lead to an equally good fit of the data. We have not taken
into account the possible effect that
newly discovered dwarf fields could become smaller and smaller 
until the new fields have a size that is so small to be economical to drill.
This effect would cap the carrying capacity at a value smaller than 
the one shown in figure  \ref{rate_of_discovery}. To address these issues, the method
described in the next section \ref{sub:Likelihood-function-for} has been used
to compute the probability of different carrying capacities.

@@

\begin{table}[h]
\caption{Likelihood estimation for the number of dwarf (1) and giant (2) fields.
For Norway, our logistic fit suggests that up to two new giant fields could be
ultimately discovered. For the U.K., the prediction is more bleak, suggesting
that the most probable scenario is that no other giant field will be found.}
\centering %
\begin{tabular}{lllll}
 & $m_{N1}$ & $n_{1}\pm\sigma_{1}$ & $m_{N2}$ & $n_{2}\pm\sigma_{2}$\tabularnewline
\midrule Norway  & $24$ & $88.4\pm10.0$ & $52$ & $56.4\pm1.6$\tabularnewline
U.K. & $162$ & $208\pm11$ & $99$ & $100\pm0.4$\tabularnewline
\bottomrule  &  &  &  & \tabularnewline
\end{tabular}\label{likelihood-results}
\end{table}

@@

As the total parameter space is too large to be sampled entirely,
a Monte Carlo technique is applied to compute the expected production
with confidence interval from future discoveries. In a nutshell, the
algorithm works as follows:
\begin{enumerate}
\item Draw a scenario (total number of fields in each size bin) based on its probability according to the likelihood function (\ref{eq:likelihood}). This is done by generating a random number $r$ between 0 and 1, and computing the scenario which is mapped to $r$ by the cumulative distribution function of all scenarios.
\item Compute the time distribution of new discoveries by fitting a logistic
curve for each size class.
\item For each discovery, generate a size and the resulting production curve
based on the size distribution and production curves of existing fields.
\item Superpose all the production curves.
\item Repeat and average over all drawn scenarios.
\end{enumerate}
The result is the expected production curve of future oil field discoveries.
The distribution of generated scenarios yields the confidence interval.

Last but not least, it has to be defined how the expected production from these future oil field discoveries is added to the extrapolated production from existing fields. The start of the simulation of new discoveries does not match up with the date of the latest production data. The reason being that the new fields (defined in section \ref{sub:Regular,-irregular-and}),
which are already discovered but did not yet enter the decay phase, are not taken into account in the simulation as their final size is not known. The only meaningful way of treating the new fields is to consider them as a discovery. Therefore, the starting point in time of the simulated production resulting from new discoveries has been choosen as the date in the past where it matches the current production from
new fields. In order words, the extrapolated production from regular and irregular fields added to the production from simulated future discoveries (which is shifted into the past as to match the production from new fields) must be equal to the current (latest available data) total production from all fields.

@@

\begin{table}[H]
\caption{Oil import (bbl/day) at a constant consumption of 1.5M bbl/day for
the U.K. and 0.22M bbl/day for Norway. The import for the E.U. and
Norway is a lower bound based on the changes in the U.K and Norway. Negative
numbers for Norway represent exports.}
\centering %
\begin{tabular}{lllcc}
 & 2014 & 2020 & 2025 & 2030\tabularnewline
\midrule Norway & $-1.23\cdot10^{6}$  & $-0.88\cdot10^{6}$  & $-0.58\cdot10^{6}$  & $-0.43\cdot10^{6}$ \tabularnewline
U.K. & $0.7\cdot10^{6}$  & $0.9\cdot10^{6}$  & $1.0\cdot10^{6}$  & $1.1\cdot10^{6}$ \tabularnewline
E.U.+Norway (lower bound) & $9.8\cdot10^{6}$  & $10.45\cdot10^{6}$  & $10.85\cdot10^{6}$  & $11.1\cdot10^{6}$ \tabularnewline
\bottomrule  &  &  &  & \tabularnewline
\end{tabular}\label{needs}
\end{table}

@@

\bibitem{Laherr=0000E8re 2002}Laherrere, Jean. \textquotedblleft{}Forecasting
Future Production from Past Discovery.\textquotedblright{} International
Journal of Global Energy Issues 18, no. 2 (2002): 218\textendash{}38.

\bibitem{Laherr=0000E8re and Sornette 1998}Laherrere, Jean, and Didier
Sornette. \textquotedblleft{}Stretched Exponential Distributions in
Nature and Economy:\textquoteleft{}fat Tails\textquoteright{} with
Characteristic Scales.\textquotedblright{} The European Physical Journal
B-Condensed Matter and Complex Systems 2, no. 4 (1998): 525\textendash{}39.

@@

@@

\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{oseberg}
\caption{Example of the time evolution of the oil production per day of
a regular field, parameterised beyond the peak in the decay
regime by the stretched exponential 
function (\ref{stretched_exponential}) with $\beta=0.66\pm0.01$ and $\tau=-55\pm1$ months,
shown with the black line.  The one standard deviation given by expression
(\ref{eq:extrapolation}) is represented by the grey band.}
\label{regular} 
\end{figure}

@@

\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{vale}
\caption{Example of a irregular field.}
\label{irregular} 
\end{figure}

@@

\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{skarv}
\caption{Example of a new field.}
\label{new} 
\end{figure}

\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{oesberg-error}
\caption{Oesberg field - Relative error defined by expression (\ref{ryjryukoik})
of the predicted total production from time $t$ indicated in the abscissa until 2014.
One can observe that the predicted future total production is over-estimated by as much
as 70\% in 1999, then under-estimated by the same amount in 2002, while the forecast
errors remain smaller than 20\% since 2004.}
\label{oesberg-error}
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\columnwidth]{extrapolating-no-discoveries}
\caption{Logistic fit of the function solution of expression (\ref{logistic})
to the number of discoveries for Norway. The discovery
rate of new oil fields is dependent on their size as explained in the main text.}
\label{rate_of_discovery} 
\end{figure}

\begin{figure}[p]
\includegraphics[width=1\columnwidth]{NO-future}
\includegraphics[width=1\columnwidth]{UK-future} 
\caption{Monte-Carlo (green upper
continuous line with standard deviation bank starting in 2014 onward)
and fit forecast based on past production data (lower line and grey one standard deviation band) for Norway
(top) and the U.K. (bottom). In both cases, the Monte-Carlo model forecasts
a significantly slower decay than the fit by taking into account that
new fields will come in production.}

\label{no_uk_2013} 
\end{figure}

\noindent 
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{NO-back-test}

\includegraphics[width=1\columnwidth]{UK-back-test} \caption{Monte-Carlo (green upper
continuous line with standard deviation bank starting in 2014 onward) and Hubbert forecast based 
on past production data up to 2008 (lower line and grey one standard deviation band)
for Norway (top) and the U.K. (bottom). The results can be
compared with the subsequent oil production (blue area). In both cases,
the Monte-Carlo methodology is more precise.}

\label{model-back-testing} 
\end{figure}

>>>>

%% LyX 2.0.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[twoside]{article}
\usepackage{mathpazo}
\renewcommand{\ttdefault}{mathpazo}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{float}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{esint}
\usepackage{url}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=false,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 1},backref=section,colorlinks=false]
 {hyperref}
\usepackage{breakurl}

@@

The idea behind the methodology is to model the future aggregate oil
production of a country by studying the production dynamics of its
individual constituents, the oil fields. The main benefit of this
approach, compared to working directly with aggregate production data,
is the possibility to forecast non-trivial oil production profiles
arising from the combination of all the individual field
dynamics. This ability reflects directly the fact that the total
production is the sum of the contributions of each individual
oil field in production. Thus modelling at the level of each field
reflects more closely the reality and is likely to be over-performing
and more reliable, as we show below.
In order to implement our approach, one must be able 1) to \textbf{extend
the oil production of each individual field} into the future and 2)
to \textbf{extrapolate the rate of discoveries} of new oil fields.

@@

The first step to predicting the future oil production of a country
is to extrapolate the future production of existing fields and to
estimate the error on this extrapolation. The data of the fields developed
in the past shows a repeating asymmetric pattern. A good example is
the Oseberg field shown in figure \ref{regular}, with a quick ramp
up once the field is being developed, and then a peak or plateau before
the oil field production starts decaying. The decay can take many different shapes
and is governed by a variety of geological and economical factors.
The goal of the fitting procedure is to capture as much the impact of
these different factors as possible.

@@

The resulting plot shown in figure \ref{rate_of_discovery} pictures
the dynamics: giant oil fields have mostly been found while the discovery
process for dwarf fields is still ongoing. The logistic growth curve
fit to the giant discoveries is well constrained, however the fit
for the dwarfs is poorly constrained. As can be seen in figure \ref{rate_of_discovery},
the carrying capacity $K$ of the logistic growth model is not well
constrained by the available data. A large spectrum of values for
$K$ can lead to an equally good fit of the data. We have not taken
into account the possible effect that
newly discovered dwarf fields could become smaller and smaller 
until the new fields have a size that is so small to be economical to drill.

There are in fact two competing effects that are likely to compensate each other.
On the one hand, figure \ref{ccdffieldsizes} shows the complementary 
cumulative distribution function (CCDF) of known oil field sizes $S$ from Norway and the UK.
Two salient properties can be observed. First, the tails 
of the distributions are well described by power laws
\begin{equation}
{\rm CCDF}(S) \sim {1 \over S^{1+\alpha}}~,
\label{ccdfsizefield}
\end{equation}
with exponent $\alpha = 1.2 \pm 0.1$ for Norway and $\alpha = 1.4 \pm 0.1$ for the UK.
The fact that the estimations of the exponents $\alpha$ are larger than $1$ implies
that the cumulative oil reserves are asymptotically controlled by the largest fields, and not the small ones
\cite{Sornettecritbook}. However, the fact that the exponents $\alpha$ are rather close
to $1$ (which is called ``Zipf law'') would make the many small oil field contributing 
significantly in total. This brings us to the second important feature exhibited by figure \ref{ccdffieldsizes},
namely the roll-overs of the CCDFs for small oil fields, likely due to an under-sampling
of the data. Indeed, as for most data sets involving broad distributions of sizes such
as oil fields, the distributions are in general incomplete for the small events due to the non
exhaustive sampling. This incompleteness raises the spectre that our extrapolations
might be grossly underestimating the large potential contributions to the total reserves
of the many small yet undiscovered small oil fields. Assuming that the power law (\ref{ccdfsizefield})
would hold for smaller fields down to size of 1 Million barrels leads to a number of 
such fields 10 to 100 times larger than presently known.

But there is another
key factor that needs to be considered, namely the fact that small oil fields
are not economically viable for exploitation, which leads to an effective truncation
in the distribution (\ref{ccdfsizefield}) relevant for the estimation of recoverable oil.
Taking the data from the NPD\cite{Norwegian Data}
providing the yearly investments broken by fields in Norway, let us consider the illustrative
case of the field GAUPE. The investment spent to 
develop its exploitation was  \$380M, while its estimated size is 
$\simeq 1.2Mb$, which corresponds approximately to a total
market value of $\$120M$ at \$100 per barrel. It is thus not a surprise that
investment to exploit this field was interrupted in 2013. 
Figure \ref{returnNorwayfield} shows the total estimated revenue divided
by investment as a function of the estimated ultimate recovery for a number
of small oil fields. It can be inferred that oil fields of sizes smaller than about 10 Mb
are not economically viable as long as the market oil price does not 
grow much higher than \$100 per barrel. This implies that, notwithstanding
the large number of unknown small oil fields, economic considerations
oblige us to neglect the small oil fields, therefore providing
a justification of our procedure. In fact, economic constraints may lead to 
cap the carrying capacity at a value smaller than 
the one shown in figure  \ref{rate_of_discovery}.

To address these issues from a more solid angle, the method
described in the next section \ref{sub:Likelihood-function-for} has been used
to compute the probability of different carrying capacities.

@@

\begin{table}[h]
\caption{Likelihood estimation for the number of dwarf (1) and giant (2) fields.
For Norway, our logistic fit suggests that up to two new giant fields could be
ultimately discovered. For the U.K., the prediction is more bleak, suggesting
that the most probable scenario is that one giant field will be found (which is most likely already discovered but classified as a new field).}
\centering %
\begin{tabular}{lllll}
 & $m_{N1}$ & $n_{1}\pm\sigma_{1}$ & $m_{N2}$ & $n_{2}\pm\sigma_{2}$\tabularnewline
\midrule Norway  & $24$ & $88.4\pm10.0$ & $52$ & $56.4\pm1.6$\tabularnewline
U.K. & $162$ & $208\pm11$ & $99$ & $100\pm0.4$\tabularnewline
\bottomrule  &  &  &  & \tabularnewline
\end{tabular}\label{likelihood-results}
\end{table}

@@

As the total parameter space is too large to be sampled entirely,
a Monte Carlo technique is applied to compute the expected production
with confidence interval from future discoveries. In a nutshell, the
algorithm works as follows:
\begin{enumerate}
\item Draw a scenario (total number of fields in each size bin) based on its probability according to the likelihood function (\ref{eq:likelihood}). This is done by generating a random number $r$ between 0 and 1, and computing the scenario that is mapped to $r$ by the cumulative distribution function of all scenarios.
\item Compute the time distribution of new discoveries by fitting a logistic
curve for each size class.
\item For each discovery, generate a size and the resulting production curve
based on the size distribution and production curves of existing fields.
\item Superpose all the production curves.
\item Repeat and average over all drawn scenarios.
\end{enumerate}
The result is the expected production curve of future oil field discoveries.
The distribution of generated scenarios yields the confidence interval.

Last but not least, it has to be defined how the expected production from these future oil field discoveries is added to the extrapolated production from existing fields. The start of the simulation of new discoveries does not match up with the date of the latest production data, the reason being that the new fields (defined in section \ref{sub:Regular,-irregular-and}), which are already discovered but did not yet enter the decay phase, are not taken into account in the simulation as their final size is not known. The only meaningful way of treating the new fields is to consider them as a discovery. Therefore, the starting point in time of the simulated production resulting from new discoveries has been choosen as the date in the past where it matches the current production from
new fields. In order words, the extrapolated production from regular and irregular fields added to the production from simulated future discoveries (which is shifted into the past as to match the production from new fields) must be equal to the current (latest available data) total production from all fields.

@@

\begin{table}[H]
\caption{Oil import (bbl/day) at a constant consumption of 1.5M bbl/day for
the U.K. and 0.22M bbl/day for Norway. The import for the E.U. and
Norway is a lower bound based on the changes in the U.K and Norway. Negative
numbers for Norway represent exports.}
\centering %
\begin{tabular}{lllcc}
 & 2014 & 2020 & 2025 & 2030\tabularnewline
\midrule Norway & $-1.23\cdot10^{6}$  & $-0.88\cdot10^{6}$  & $-0.58\cdot10^{6}$  & $-0.43\cdot10^{6}$ \tabularnewline
U.K. & $0.7\cdot10^{6}$  & $0.9\cdot10^{6}$  & $1.0\cdot10^{6}$  & $1.1\cdot10^{6}$ \tabularnewline
E.U.+Norway & $9.8\cdot10^{6}$  & $10.45\cdot10^{6}$  & $10.85\cdot10^{6}$  & $11.1\cdot10^{6}$ \tabularnewline
\bottomrule  &  &  &  & \tabularnewline
\end{tabular}\label{needs}
\end{table}

@@

\bibitem{Laherr=0000E8re 2002}Laherr\`{e}re, Jean. \textquotedblleft{}Forecasting
Future Production from Past Discovery.\textquotedblright{} International
Journal of Global Energy Issues 18, no. 2 (2002): 218\textendash{}38.

\bibitem{Laherr=0000E8re and Sornette 1998}Laherr\`{e}re, Jean, and Didier
Sornette. \textquotedblleft{}Stretched Exponential Distributions in
Nature and Economy:\textquoteleft{}fat Tails\textquoteright{} with
Characteristic Scales.\textquotedblright{} The European Physical Journal
B-Condensed Matter and Complex Systems 2, no. 4 (1998): 525\textendash{}39.

@@

\bibitem{Sornettecritbook} Sornette, D.,
``Critical Phenomena in Natural Sciences, Chaos, Fractals, Self-organization and Disorder: Concepts and Tools'',
2nd ed., Springer Series in Synergetics, Heidelberg, 2004.

@@

%FIGURE 1
\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{oseberg}
\caption{Example of the time evolution of the oil production per day of
a regular field, parameterised beyond the peak in the decay
regime by the stretched exponential 
function (\ref{stretched_exponential}) with $\beta=0.66\pm0.01$ and $\tau=-55\pm1$ months,
shown with the black line.  The one standard deviation given by expression
(\ref{eq:extrapolation}) is represented by the grey band.}
\label{regular} 
\end{figure}

%FIGURE 2
\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{vale}
\caption{Example of a irregular field.}
\label{irregular} 
\end{figure}

@@

%FIGURE 3
\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{skarv}
\caption{Example of a new field.}
\label{new} 
\end{figure}

@@

%FIGURE 4
\noindent 
\begin{figure}[H]
\includegraphics[width=1\columnwidth]{oesberg-error}
\caption{Oesberg field - Relative error defined by expression (\ref{ryjryukoik})
of the predicted total production from time $t$ indicated in the abscissa until 2014.
One can observe that the predicted future total production is over-estimated by as much
as 70\% in 1999, then under-estimated by the same amount in 2002, while the forecast
errors remain smaller than 20\% since 2004.}
\label{oesberg-error}
\end{figure}

%FIGURE 5
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{extrapolating-no-discoveries}
\caption{Logistic fit of the function solution of expression (\ref{logistic})
to the number of discoveries for Norway. The discovery
rate of new oil fields is dependent on their size as explained in the main text.}
\label{rate_of_discovery} 
\end{figure}

%FIGURE 6
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{ccdf-field-size}
\caption{Complementary 
cumulative distribution function (CCDF) of known oil field sizes $S$ from Norway and the UK.
The two dashes lines visualise the power law behaviour (\ref{ccdfsizefield}) of the tail
of the distributions. }
\label{ccdffieldsizes}
\end{figure}

%FIGURE 7
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{no-returns-size}
\caption{Total estimated revenue divided by investment as a function of the estimated 
ultimate recovery for a number of small oil fields. }
\label{returnNorwayfield}
\end{figure}

%FIGURE 8
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{NO-future}
\includegraphics[width=1\columnwidth]{UK-future} 
\caption{Monte-Carlo (green upper
continuous line with standard deviation bank starting in 2014 onward)
and fit forecast based on past production data (lower line and grey one standard deviation band) for Norway
(top) and the U.K. (bottom). In both cases, the Monte-Carlo model forecasts
a significantly slower decay than the fit by taking into account that
new fields will come in production.}
\label{no_uk_2013} 
\end{figure}

%FIGURE 9
\noindent 
\begin{figure}[p]
\includegraphics[width=1\columnwidth]{NO-back-test}
\includegraphics[width=1\columnwidth]{UK-back-test} \caption{Monte-Carlo (green upper
continuous line with standard deviation bank starting in 2014 onward) and Hubbert forecast based 
on past production data up to 2008 (lower line and grey one standard deviation band)
for Norway (top) and the U.K. (bottom). The results can be
compared with the subsequent oil production (blue area). In both cases,
the Monte-Carlo methodology is more precise.}
\label{model-back-testing} 
\end{figure}
